---
title: "R 4.0 installation guide"
author: "Antoine Languillaume"
date: "6/2/2020"
output: 
  github_document:
    # number_sections: true
---

# Why update to R 4.0 ?



## R set-up good practices on windows WUR Sytems

On WUR windows system, I believe it is good practice to avoid installing software in `C:/Program Files/`.
For the simple reason that has a lambda user you don't have writing permission on that folder. Thus, you won't be able to modify the configuration files of your different programs. A good work-around is to create a folder `C:/MyPrograms/` where you will manually install software upon which you need to have finer control on. 

A snapshot of my own `C:/MyPrograms/` directory:

```{r}
fs::dir_tree("C:/MyPrograms/", recurse = FALSE)
```

__IMPORTANT__: To be able to install software yourself you need to have __power-user__ rights. This enables you to run program installers with administrative rights simply by right-clicking on the .exe installer and selecting __WUR - Run with administrative rights__. If you have a WUR laptop it should be the default option, if not you will have to call IT, so that they will grant you that privilege.


## Update R


### Download R for windows

[Download R 4.0](https://cran.r-project.org/bin/windows/base/})

### Install R

Execute the installer with the __WUR - Run with administrative rights__ option and make sure to install R in `C:/MyPrograms/R/`.



## Update Rstudio


### Uninstall your previous version of Rstudio:

Just to be sure. Type `WUR - install applications` in the start-up menu and click on the corresponding item. Then navigate to Rstudio in the applications list, right click on it and uninstall it.

### Download Rstudio for windows

[Download Rstudio 1.3](https://rstudio.com/products/rstudio/download/#download)

### Install Rstudio

Execute the installer with the __WUR - Run with administrative rights__ option and make sure to install R in `C:/MyPrograms/Rstudio/`.



## Update Rtools

Rtools are the set of command line tools that are used to build R and R packages from source, something you have to do from time to time to get the latest version of a package. It is easy to install, you never have to use it directly and it completes your setup. You'd better do that now as you are taking some time to upgrade your setup rather than later when you will be trying to quickly run an analysis to meet some important deadlines...

### Uninstall your previous version of Rtools:

If you already have Rtools installed, It should be version 3.5 or so. Type `WUR - install applications` in the start-up menu and click on the corresponding item. Then navigate to Rtools in the applications list, right click on it and uninstall it.


### Download Rtools

[Downlad Rtools 4.0](https://cran.r-project.org/bin/windows/Rtools/)

Download the version suited for your system (32 or 64-bit). By now, you should now have a 64-bit computer.
If you have no clue about that go to `Control Panel\System and Security\System` or simply type `sessionInfo()` in the R console.

### Install Rtools

Execute the installer with the __WUR - Run with administrative rights__ option and make sure to install R in `C:/MyPrograms/Rtools/`.

### Putting Rtools on the PATH

After installation is complete, you need to perform one more step: you need to put the location Rtools on the your computer PATH which is basically a system wide address book that allows program to know where to find each other.
To do so we are going to edit a R start up file called `Renviron.site`. Don't worry it is REALLY simple.

At the R console type the following command:

```{r, eval = FALSE}
file.edit(R.home("etc/Renviron.site"))
```

Add the following lines, save and close the `Renviron.site` file.

```{r, eval = FALSE}
# Set Rtools path
PATH="${RTOOLS40_HOME}\\usr\\bin;${PATH}"
```

Now restart R, type the following command in the R console. That should return something along those lines:

```{r}
Sys.which("make")
```

If this works, you can try to install an R package from source:

```{r, eval = FALSE}
install.packages("jsonlite", type = "source")
```

If this succeeds, you’re good to go!


## Reinstall your packages

### General advice

Your R packages are gather in a special folder called a __library__. By default on  WUR windows system, that library is created in `M:/Documents/R/win-library/` which means it goes dangling somewhere on the WUR server. This clearly suboptimal as you have your packages stored on a different network then R itself. It is a bit like having your workshop in your basement and all your tools in the attic, not really efficient. Moreover, if you don't fix that you will get an error each time you execute a command (something like :`\\\WURNET\\...`).
So for thing to as smoothly as possible, I usually have my libraries in `C:/MyPrograms/R-libraries/`.

#### Setting up R libraries in the proper locaton:

First, create a folder called: `C:/MyPrograms/R-libraries/`.

Then you will have to tell R upon start-up where your packages actually are. This is fairly simple, just like we did with Rtools, we are gonna edit the Renviron.site file once again:

At the R console type the following command:

```{r, eval = FALSE}
file.edit(R.home("etc/Renviron.site"))
```

Add the following lines, save and close the `Renviron.site` file.

```{r, eval = FALSE}
## Set location of user-installed packages
R_LIBS_USER=${R_LIBS_USER-'C:/MyPrograms/R-libraries/4.0'}
```


### Updating your packages

R 4.0 is a major release so it comes with brand you internals. In non-jargon it means that R and R packages are built in a slightly different way, which unfortunately forces the users to re-install all their packages. To do that:

*	Make sure your CRAN mirror (the server where you download your packages) is set to the closet location: in RStudio, Tools > Global Options > Packages > primary CRAN repository. For us at WUR the closest is `Belgium (Ghent) [https] - Ghent University Library`. If you are working from a different place, please pick the mirror closest to where you are.

*	Find your current R library, in the R console type:

```{r, eval = FALSE}
.libPaths()
```

Something along those lines should appear: `"C:/Users/Someone/Documents/win-library/4.0"`.

*	Close to that library should be your former R library with all the packages previously installed, something like: `"C:/Users/Juani/Documents/win-library/3.6"`

*	Still in the console now type: 

__MAKE SURE YOU REPLACE THE PATH WITH YOURS !__

```{r, eval = FALSE}
install.packages(list.files ("C:/Users/Juani/Documents/win-library/3.6")) 
```

This little hack should save you time. Just wait for everything to be installed.


And eh voilà ! You should be ready with a sound and safe R 4.0 setup !


